MassWateR: Improving quality control, analysis, and sharing of water quality data

### 2023-09-18 revision/review-response.qmd

Though one of the reviewers recommended a complete rejection, I found that this study could be of benefit to the community especially in the industry 4.0 era. However, I request that you take into consideration the following while revising your manuscript.

1.  You should describe explicitly how the package provides a better alternative to all the other existing packages. Where applicable, cite the weaknesses of the existing packages with supporting citations.

**Response**: Thank you for this comment. As noted in the original submission on line 88, there are no other R packages that we are aware of that provide similar functionality. As such, there is no basis of comparison of MassWateR to determine relative strengths/weakness with other packages. However, we have added text describing other R packages that are somewhat related and have included a description of software packages outside of the R environment to provide a more comprehensive discussion of alternatives. The following additions were made:

"The *TADA* R package [@Mullin22] is also currently under development as a resource for compiling and evaluating data from the WQP. *TADA* is similar to *dataRetrieval* in that it can used for importing data, but the package is also expected to provide more comprehensive methods for cleaning, filtering, and processing data using the rich qualifier codes provided by WQP."

"Other software platforms outside of the R environment provide various services for quality control of water quality data. For example, the [Aquarius](https://aquaticinformatics.com/products/aquarius-environmental-water-data-management/) software environment is a proprietary resource for managing hydrologic and water quality data and is used by several private and public institutions. Functionality is provided to pre-process and synthesize multiple data streams, manually correct erroneous values, and to visualize results for decision-making. A graphical user interface is provided to access the various features of the software, as compared to a programmatic approach for building custom routines. Moreover, the software is not open-source, which implicitly limits its development to a core set of maintainers and is not freely accessible to the broader community [e.g., @schrape2017a]. Similar platforms are available from companies that manufacture data loggers (e.g., [YSI](https://www.ysi.com/), [HOBO](https://www.onsetcomp.com/products/software/hoboware)), all of which are specific to the monitoring equipment and not broadly transferable. Alternative publicly funded software and data services are provided by the Consortium of Universities for the Advancement of Hydrologic Science ([CUAHSI](https://www.cuahsi.org/data-services/solutions)). Services provided by CUAHSI include data discovery, archiving, cloud computing, and analysis, with many of the services available in open source environments (e.g., [Python](https://www.python.org/)). However, the services provided by CUAHSI are meant to address a variety of different resources and use cases, none of which are immediately related to quality control reporting and data submission to public institutions. As such, specific software solutions to address these needs and that leverage existing tools available in a rich open-source environment are clearly needed."

@@

2.  Make appropriate citations to literature sources. For example, in L26 and L29, you talk about an Act and an EU framework whose sources are not cited. The same applies to L39-40, where you argue out why this communication could be important but without any supporting citation(s). Similar omissions are evident in the DISCUSSION where you should make appropriate citation of the "community of practice established" and any outcomes (L521-537, and so forth).

**Response**: We have added citations where needed.

Line 26 includes a link to a description of the US Clean Water Act ([33 USC § 1251 et seq.](https://uscode.house.gov/view.xhtml?path=/prelim@title33/chapter26&edition=prelim), note that there is no citation provided in the references because this is legislative code).

Line 29 includes a citation to the European Water Framework Directive [Directive 2000/60/EC of the European Parliament and of the Council of 23 October 2000, @pollard1998]

Line 40 includes the following citations: @kinzelman2003, @Wagner06, @zhang2006.

Lines 521-547 include ciitations:

@@

3.  L87-99 appears to should have been part of the introduction, highlighting what the MassWateR brings onboard.

**Response**: This paragraph and the additions shown above were moved to the introduction.

4.  Figure 1 could best be presented as a flow/block diagram.

**Response**: Figure 1 has been reconfigured as a flow diagram, although emphasis remains in the text that a user can engage with the package using one to any of the steps in the diagram.

@@

The availability of reliable monitoring data is the basis of scientific research. This study collected discrete surface water quality data generated by MassWateR software package and analyzed the trend of parameters. The study has some fundamental significance. However, this paper is more like a technical manual than a scientific research paper. Therefore, I would recommend the authors to submit the paper to a technical journal rather than a research-oriented journal.

**Response**: Thank you for taking the time to review our paper. Based on the opinion of the academic editor, this paper is considered to be of sufficient interest to the readership of the journal. Additionally, the paper is not written as a research article and was submitted as a software manuscript under the acceptable formats considered by the journal (see [here](https://journals.plos.org/plosone/s/submission-guidelines#loc-methods-software-databases-and-tools)). The format of these submissions is notably different from those of conventional research articles. In particular, papers describing software must demonstrate utility, validation, and availability. Please see our comment to the academic editor above regarding these requirements.

In addition, the transferability of the developed tool needs to be demonstrated and analyzed, which is directly related to its application value. The authors should give concrete examples in this area;

**Response**: Thank you for this suggestion. In our original submission, we provided substantial text describing how MassWateR can be used in more generic applications, i.e., its transferability. This information was provided in the discussion on lines 538-556, where specific considerations were noted for how users could apply the software to novel datasets. Truthfully, we are not yet aware of specific applications of MassWateR outside of our core user base, but our intention of publishing this work in the primary literature was to reach a broader audience and thereby promote transferibility of the software. We are confident that the software can be broadly applied given the relatively minor limitations for its use outside of Massachusetts. Additionally, the software has been downloaded over 4000 times from CRAN as of this writing, which suggests many others have found the software transferable.

### 2023-09-18 manu-draft.qmd
adding references

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act ([33 USC § 1251 et seq.](https://uscode.house.gov/view.xhtml?path=/prelim@title33/chapter26&edition=prelim)) with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters. Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union [Directive 2000/60/EC of the European Parliament and of the Council of 23 October 2000, @pollard1998]. Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances identified from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies. Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements.

@@

To our knowledge, there are no existing R packages on the Comprehensive R Archive Network ([CRAN](https://cran.r-project.org/)) that can be used to facilitate QC of water quality data, nor are any available that facilitate submission to existing databases. However, there are several that can be used to retrieve and analyze data from existing sources (see the CRAN [Hydrology](https://cran.r-project.org/web/views/Hydrology.html) Task View). In particular, the *dataRetrieval* R package [@DeCicco22] has been used widely to retrieve data from the WQP, which is the interface for accessing data submitted using WQX. This package leverages a robust API to query existing water quality data in standardized format provided by the WQP. The *TADA* R package [@Mullin22] is also currently under development as a resource for for compiling and evaluating data from the WQP. *TADA* is similar to *dataRetrieval* in that it can used for importing data, but the package is also expected to provide more comprehensive methods for cleaning, filtering, and processing data using the rich qualifier codes provided by WQP. As such, data retrieval using existing web services is much simpler than data submission, as data formatting requirements do not apply when retrieving data. Developing a robust tool that can facilitate the upload of data to WQX, in addition to streamlining QC processes, would further the value of packages like *dataRetrieval* and *TADA* by increasing the amount of data that can be accessed through the WQP. The *MassWateR* package was developed to provide this benefit.

Other software platforms outside of the R environment provide various services for quality control of water quality data. The [Aquarius](https://aquaticinformatics.com/products/aquarius-environmental-water-data-management/) software environment is a proprietary resource for managing hydrologic and water quality data and is used by several private and public institutions. Functionality is provided to pre-process and synthesize multiple data streams, manually correct erroneous values, and to visualize results for decision-making. A graphical user interface is provided to access the various features of the software as compared to a programmatic approach for building custom routines. Moreover, the software is not open-source, which implicitly limits its development to a core set of maintainers and is not freely accessible to the broader community [e.g., @schrape2017a]. Similar platforms are available from companies that manufacture data loggers (e.g., [YSI](https://www.ysi.com/), [HOBO](https://www.onsetcomp.com/products/software/hoboware)), all of which are specific to the monitoring equipment and not broadly transferable. Alternative publicly funded software and data services are provided by the Consortium of Universities for the Advancement of Hydrologic Science ([CUAHSI](https://www.cuahsi.org/data-services/solutions)). Services provided by CUAHSI include data discovery, archiving, cloud computing, and analysis, with many of the services available in open source environments (e.g., [Python](https://www.python.org/)). However, the services provided by CUAHSI are meant to address a variety of different resources and use cases, none of which are immediately related to quality control reporting and data submission to public institutions. As such, Specific software solutions to address these needs and that leverage existing tools available in a rich open-source environment are clearly needed.

@@

![Workflow demonstrating how a user could engage with the *MassWateR* package. WQX: Water Quality Exchange; QC: Quality Control.](images/workflow.tif){#fig-workflow width="100%" fig-alt="Workflow diagram showing four ways to engage with *MassWateR*"}

>>>>

Water quality measurements provide the foundation of environmental monitoring programs designed to protect or restore aquatic resources. In the United States, these programs are broadly guided by the federal Clean Water Act ([33 USC § 1251 et seq.](https://uscode.house.gov/view.xhtml?path=/prelim@title33/chapter26&edition=prelim)) with the singular goal of restoring and maintaining the chemical, physical, and biological integrity of the nation's surface waters. Similarly, the Water Framework Directive provides a legislative foundation for the protection of aquatic resources in member states of the European Union (Directive 2000/60/EC of the European Parliament and of the Council of 23 October 2000) [@pollard1998]. Numeric standards that define critical thresholds for protecting recreational, aquatic life, industrial, navigational, and consumptive uses of the resource are often established by government agencies, such that exceedances identified from water quality measurements require additional regulatory action to ensure compliance. These standards and other regulatory assessments as applied at the state-level use information from long-term monitoring datasets [@Schiff16; @Tango13], or data collected *ad hoc* from multiple assessment endpoints [@Stein09; @Behmel16; @Kumpel20], where the former is atypical for most surface water bodies. Many state or regional institutions that assess water quality rely on decentralized data sources, often combining datasets from local watershed groups or participatory science programs rather than a single database that contains adequate coverage of all areas of interest [@Buytaert14; @Kelly22]. Use of these monitoring data in a regulatory context is not possible unless standard operating procedures are adopted and the data fulfill quality control (QC) requirements [e.g., @kinzelman2003; @Wagner06; @zhang2006].

@@

To our knowledge, there are no existing R packages on the Comprehensive R Archive Network ([CRAN](https://cran.r-project.org/)) that can be used to facilitate QC of water quality data, nor are any available that facilitate submission to existing databases. However, there are several that can be used to retrieve and analyze data from existing sources (see the CRAN [Hydrology](https://cran.r-project.org/web/views/Hydrology.html) Task View). In particular, the *dataRetrieval* R package [@DeCicco22] has been used widely to retrieve data from the WQP, which is the interface for accessing data submitted using WQX. This package leverages a robust API to query existing water quality data in standardized format provided by the WQP. The *TADA* R package [@Mullin22] is also currently under development as a resource for compiling and evaluating data from the WQP. *TADA* is similar to *dataRetrieval* in that it can used for importing data, but the package is also expected to provide more comprehensive methods for cleaning, filtering, and processing data using the rich qualifier codes provided by WQP. As such, data retrieval using existing web services is much simpler than data submission, as data formatting requirements do not apply when retrieving data. Developing a robust tool that can facilitate the upload of data to WQX, in addition to streamlining QC processes, would further the value of packages like *dataRetrieval* and *TADA* by increasing the amount of data that can be accessed through the WQP. The *MassWateR* package was developed to provide this benefit.

Other software platforms outside of the R environment provide various services for quality control of water quality data. For example, the [Aquarius](https://aquaticinformatics.com/products/aquarius-environmental-water-data-management/) software environment is a proprietary resource for managing hydrologic and water quality data and is used by several private and public institutions. Functionality is provided to pre-process and synthesize multiple data streams, manually correct erroneous values, and to visualize results for decision-making. A graphical user interface is provided to access the various features of the software, as compared to a programmatic approach for building custom routines. Moreover, the software is not open-source, which implicitly limits its development to a core set of maintainers and is not freely accessible to the broader community [e.g., @schrape2017a]. Similar platforms are available from companies that manufacture data loggers (e.g., [YSI](https://www.ysi.com/), [HOBO](https://www.onsetcomp.com/products/software/hoboware)), all of which are specific to the monitoring equipment and not broadly transferable. Alternative publicly funded software and data services are provided by the Consortium of Universities for the Advancement of Hydrologic Science ([CUAHSI](https://www.cuahsi.org/data-services/solutions)). Services provided by CUAHSI include data discovery, archiving, cloud computing, and analysis, with many of the services available in open source environments (e.g., [Python](https://www.python.org/)). However, the services provided by CUAHSI are meant to address a variety of different resources and use cases, none of which are immediately related to quality control reporting and data submission to public institutions. As such, specific software solutions to address these needs and that leverage existing tools available in a rich open-source environment are clearly needed.

@@

![Workflow demonstrating how a user could engage with the *MassWateR* package. DQO: Data Quality Objectives; QC: Quality Control; WQX: Water Quality Exchange.](images/workflow.tif){#fig-workflow width="100%" fig-alt="Workflow diagram showing four ways to engage with *MassWateR*"}
